MSstatsPTM statistical relative quantification of post-translational modifications in bottom-up proteomic experiments

### 2022-03-28 main/MSstatsPTM_Main.tex
rewrote background section

Each simulation was generated with random biological variation. Specifically, let $Y_{ibc}$ denote the abundance of peptide $i$ in biological replicate $b$ of condition $c$. The simulated peptide abundance  $Z_{ibc}$ was generated as $Z_{ibc} = Y_{ibc} + \epsilon_{ibc}$ where $\epsilon_{ibc}\sim^{iid}N(0,\sigma^2)$. The simulations were generated with one of two values $\sigma^2 = \{.2, .3\}$. The values for $\sigma^2$ were motivated \todo{motivated how specifically? corresponded to median variances?}  by the biological experiments in this manuscript.
Further details are in Supplementary Sec. \ref{supp-sec:dataset1}.

@@

\subsection*{Dataset 5 : Mouse - Phosphorylation - 2mix-TMT time series}
\label{sec:exp_proc_dataset5}
In this experiment 8 biological replicates were allocated to 2 TMT mixtures in an unbalanced \todo{time course} design. More details are in Supplementary Sec. \ref{supp-sec:shigella}.

@@

The proteome was quantified in 11-plex isobaric multiplexing via tandem mass tagging (TMT) in combination with LC-MS/MS. The six conditions were split between 11 channels, and as the result the experiment was unbalanced. Each mixture contained two replicates per early and late WT and KO conditions. Mixture one contained one replicate of uninfected WT and two replicates of uninfected KO. Mixture two contained one replicate of uninfected KO and two uninfected WT. \todo{was there a reference channel?}

@@

In this experiment 8 biological replicates were split between 4 conditions in a balanced \todo{group comparison} design. Further details in Suppler Sec \ref{supp-sec:usp30}.

@@

Figure \ref{fig:data-structure}(a) schematically illustrates this data structure for one protein in a simplified experiment with $I=2$ conditions. \todo{Describe the population parameters in the figure (quantities of interest}.  The data are collected in $J=2$ \todo{biological?} replicate runs \todo{describe how the data are representative of the population parameters of interest}
The number of quantified features vary across replicate LC-MS/MS runs of the same sample, and across conditions. \todo{There may be outliers}. Some spectral features can be missing.

@@

\noindent {\bf Two-sample $t$-test} \todo{When describing the existing methods, could you refer to Fig 1? Here is the suggested order: (1) What is the null hypothesis? Do these methods test the same null hypothesis? (If not, it may be a good idea to add this null hypothesis to the figure for contrast) (2) What are the data (how summarized prior to the testing); (3) what is the model assumptions? (4) how inference is performed} Many investigations perform differential abundance analysis of PTMs using two-sample $t$-test or its extensions.

\todo{These sentences were combined from different places. Could you edit?} The approach takes as input intensities of individual features from modified peptides, or intensity ratios of modified and unmodified peptide features, and compares the mean abundance of a PTM site from one condition to another~\cite{Schwammle2015}\cite{THOMAS2020}. Two-sample $t$-test is based on the null hypothesis that there is no difference in mean PTM abundance between Conditions $i$ and $i^*$. The abundance in each run is taken as input and is often estimated by sum of peak intensities. For example, the log-abundance estimate for the PTM in Run $j$ of Condition $i$ is given by
\[
\log \left( \sum_{k=1}^{K} 2^{y_{ijk}^{\ast}} \right).
\]
For adjustment with respect to unmodified peptides, the estimate of PTM abundance is divided by the protein abundance estimate, 
and the $t$-test for the adjusted PTM abundance on log scale takes as input the difference of their log-estimates.
%which is equivalent to take the difference of their log abundances. 
The quantity is denoted by $d_{ij}$ and is given by
\[
d_{ij} = \log \left( \sum_{k=1}^{K} 2^{y_{ijk}^{\ast}} \right) - \log \left( \sum_{l=1}^{L} 2^{y_{ijl}} \right).
\]
Alternatively, run-level summarization can also be used. The difference between the means of PTM abundance in Conditions $i$ and $i'$ are estimated as
\[
\hat{\Delta} = \frac{1}{J} d_{i+}^{\ast} - \frac{1}{J} d_{i'+},
\]
where $d_{i+}^{\ast} = \sum_{j=1}^{J} d_{ij}^{\ast}$ and the test statistic for the $t$-test is given by $\hat{\Delta} / \mathrm{SE}(\hat{\Delta})$. The statistical significance of the difference is determined by comparing the test statistic against the $t$ distribution, with degrees of freedom $df=2J-2$ in balanced designs.

\medskip \noindent {\bf Limma} \todo{Could you fine-tune, using the same order as above for Limma? I.e., null hypothesis, the data, the model, the inference. I think the text can be made more compact, and a bit more technical, by directly referring to either Fig 1 or the equations above} Modifications of the $t$-test such as moderated $t$-test with Limma were also proposed~\cite{Ritchie_15a}\cite{Zhu}. While simple, these approaches do not fully account for the sources of variations, and are not directly applicable to experiments with complex designs, e.g., comparisons of multiple conditions, acquisition in multiple batches, etc. Additionally, while these approaches can be applied to experiments studying PTMs, there is not a self contained, straight forward implementation of the methods, making application challenging.

Limma uses linear models to test the null hypothesis that there is no difference in mean PTM abundances between conditions. It leverages Empirical Bayes moderation to share pooled variance information across individual modification models and moderate the individual residual variances. Using a linear model allows Limma to share variance information across conditions, providing a more accurate estimate~\cite{Ritchie_15a}.

With respect to PTM analysis, feature level summarization and adjustment for changes in the unmodified peptides are performed in the same way as $t$-test. After fitting a linear model, the final variance is moderated using the global variance across all modified peptides. This is done as $s_{i} = P(v_i| v_\Theta)$ where $v_i$ is the resulting variance when modeling PTM $i$ and $v_\Theta$ is the global variance.

@@

\todo{Could you also refer to Figure 1? Essentially, MSstats only analyzes a part of the figure. Refer to the null hypothesis. Then describe the summarization (including AFT etc), and the model for he summarized data (the fact that different designs require different models - this is your argument for the limitation of the methods above.)} The proposed approach leverages the large amount of work done to create $MSstats$~\cite{Choi:2014}. $MSstats$ takes as input a list of log-transformed intensities of spectral features which are used to characterize the identified protein. For each protein, the feature log-intensities are expressed using a linear mixed model in consideration of the effects of condition, run, feature and interaction between run and feature. The model parameters are estimated using a split-plot approach, where the feature log-intensities are first summarized into a single value per protein per run in the subplot model, and the protein-level summaries are then used for the inference of the protein abundance. This approach allows the method to be extended to cases where the experimental design is unbalanced and where additional sources of variation are present. In the protein-level summarization, Tukey's median polish (TMP), a simple and robust procedure, is applied to iteratively fit a two-way additive model with the effects of run and feature, which in turn summarizes the log-intensities for each protein~\cite{Tukey:1977}. After summarization, the inference of the protein abundance in each condition is carried through fitting a model based on the family of linear mixed-effects models \cite{Bolker2009} \cite{Faraway:2006}. We leverage this statistical modeling and quantification workflow and apply it to PTMs, in addition to unmodified proteins.

@@

@@

We propose a statistical method for detecting changes in PTMs while removing confounding with changes in global protein abundance. For simplicity, we discuss the method in the special case of a balanced design with multiple conditions and one technical replicate. However, in  practice the experimental design can be unbalanced and variation can come from multiple sources, as shown in the next section.

\todo{Put it in the context of the figure again, and also in the context of the existing methods in Background - how is it similar to the existing methods, and how different} The null hypothesis states that there is no difference in log-abundance of the PTM site between conditions, adjusted with respect to protein abundance. The expected values of log-abundances of the PTM and protein in Condition $i$ are denoted by $\mu_{i}^{\ast}$ and $\mu_{i}$ respectively. The feature and run level summarization \todo{Summarized how?} for condition $i$ is denoted as $\hat{y}_{i+}^{\ast}$ for the modified peptide and $\hat{y}_{i+}$ for the unmodified peptide. For two conditions $i$ and $i'$ the hypothesis test is formulated as follows:

\begin{equation}
\begin{aligned}
H_{0}: \Delta = ( \mu_{i}^{\ast} - \mu_{i'}^{\ast} ) - ( \mu_{i} - \mu_{i'} ) = 0 \\
H_{a}: \Delta = ( \mu_{i}^{\ast} - \mu_{i'}^{\ast} ) - ( \mu_{i} - \mu_{i'} ) \neq 0
\end{aligned}
\label{eq_1}
\end{equation}

Specifically, the adjusted difference is given by the difference in log-abundance of the PTM site, subtracted by the difference in log-abundance of the underlying protein, which is equivalent to the log of the ratio of PTM abundance difference to protein abundance difference.

\begin{equation}
\hat{\Delta} = \left[ \frac{1}{J} \left( \hat{y}_{i+}^{\ast} - \hat{y}_{i'+}^{\ast} \right) \right] - \left[ \frac{1}{J} \left( \hat{y}_{i+} - \hat{y}_{i'+} \right) \right]
\end{equation}

The estimate of the adjusted difference and the standard error (SE) of the estimate are obtained by combining the difference estimates and the associated SEs from both counterparts.

\begin{equation}
SE(\hat{\Delta}) = \left[ \left( \hat{\sigma}_{\gamma^{\ast}}^{2} + \hat{\sigma}_{\gamma}^{2} \right) \right]^{1/2}
\end{equation}

@@

\begin{equation}
\left( \hat{\sigma}_{\gamma^{\ast}}^{2} + \hat{\sigma}_{\gamma}^{2} \right)^2 \bigg/
\left( \frac{\hat{\sigma}_{\gamma^{\ast}}^{4}}{\mathrm{df}(\gamma^{\ast})} + \frac{\hat{\sigma}_{\gamma}^{4}}{ \mathrm{df}(\gamma)} \right)
\end{equation}

@@

@@

The statistical modeling approaches discussed above can be extended to complex designs including experiments with additional sources of variation and unbalanced designs. In this case our null hypothesis and Equation 1 are unchanged.

Model inference is done in the same way to how MSstats~\cite{Choi:2014} and MSstatsTMT~\cite{Huang:2020} target the global protein. In cases where the design is unbalanced, a linear mixed effects model is fit, which takes into account all potential sources of variation, and restricted maximum likelihood (RML) is used to estimate the parameters of each model. Once the parameters are estimated, we can combine the models using modified versions of Equations 2, 3, and 4.

The log-fold change in the adjusted PTM abundance, $\Delta$, is now estimated by RML, and Equation 2 becomes 
\begin{equation}
\hat{\Delta} = (\hat{\mu}_{RML_i}^{\ast} - \hat{\mu}_{RML_{i'}}^{\ast}) - (\hat{\mu}_{RML_i} - \hat{\mu}_{RML_{i'}})
\end{equation}

@@

The proposed methods are implemented in the open source R package $MSstatsPTM$, available on Bioconductor. $MSstatsPTM$ includes converters for multiple spectral processing tools, including MaxQuant, Progenesis, and Spectronaut. The converters take as input the raw data from the tool, identify the modification site for modified peptide, and put the data into the correct format for analysis in $MSstatsPTM$. Conversion is done separately for the modified and global profiling runs. If the global profiling run is not available, the package can still analyze the modified run, but will do so without adjusting for changes in unmodified protein abundance. Specifically, if a global profiling run is not available the null and alternative hypothesis in Equation \ref{eq_1} will reduce to:

\begin{equation}
\begin{aligned}
H_{0}: \Delta = \mu_{i}^{\ast} - \mu_{i} = 0 \\
H_{a}: \Delta = \mu_{i}^{\ast} - \mu_{i} \neq 0
\end{aligned}
\label{eq_9}
\end{equation}

Which amounts to running the methods seen in MSstats and MSstatsTMT on peptide level data \cite{Choi:2014} \cite{Huang:2020}.

After using the converters, the next step is peptide/protein summarization and missing value imputation. In the methods sections it was assumed that all the input data was summarized; in the implementation we need to run a specific summarization step to make this happen. For the modified run, the package summarizes features, PSMs, which include the same modification together. Features with multiple modification sites are not included with single site features and are summarized separately. For the global profiling run all unmodified features from the same protein are summarized together, up to the protein level. Additionally, the summarization includes global median normalization and normalizes between MS runs. The package uses an AFT model to impute missing values, although this step is optional.

@@

The package relies on functionalities from the R packages $MSstats$ \cite{Choi:2014} and $MSstatsTMT$ \cite{Huang:2020}. The statistical modeling relies on the functionality from the R packages $lme4$ \cite{Bates2015} and $lmerTest$ \cite{Kuznetsova2017}. An overview of the steps of the package are illustrated in Figure \ref{fig:msstatsptm_design}.

@@

In this simulation the proposed method outperformed $t$-test and Limma. Changes in unmodified protein level still needed to be accounted for to control the FDR. Once controlling for changes in the unmodified protein, the proposed method outperformed the other methods, as seen in Figure \ref{fig:sim2_acc}. The proposed method calibrated model accuracy and recall rate well, even when the number of replicates were low. Additionally, when comparing the fold change estimation across all modified peptides, the proposed method showed a tighter distribution of estimated fold changes around the true fold change, as compared to $t$-test and Limma. Specifically, the inner quartile range of the estimated fold change for the proposed method was on average 10.4\% smaller than $t$-test and 21.8\% smaller than Limma. This fold change comparison can be seen in Supplementary Figure \ref{supp-fig:fc_boxplot}.

@@

In this experiment all models incorrectly estimated the fold change of the modified spike-in peptides before adjusting for changes in global protein abundance. After adjustment, the spike-in peptides' fold change was generally in line with expectation (Figure \ref{fig:spikein_boxplot}), however the distribution of estimated fold changes was visibly wider in all methods. Of the three approaches, the proposed method showed the tightest distribution around the true log fold change. Comparing the inner quartile range (IQR) of the spike-in peptide's log fold change, the proposed method's IQR was 36.78\% smaller than Limma's and 32.98\% smaller than $t$-test's. This means that the proposed method's fold change estimate was more consistently correct, when compared to Limma and $t$-test.

In Figure \ref{fig:spikein_prop_volcano} we can clearly see the red labeled spike-in peptides do not follow the expected fold change before adjusting for changes in global protein abundance. After adjustment, the estimated fold change was more in line with expectation. Additionally, the background peptides serving as the null model, show many false positives before adjustment was made. After adjustment the results improved and the number of false positives decreased significantly. Specifically, for the proposed method, the number of false positives went from 20.88\% to 1.84\% of all true negatives after adjustment was applied. Using both the proposed method and Limma, the red labeled spike-in peptides were more inline with expected fold change after the adjustment was applied. However, using Limma resulted in the majority of the differential spike-in peptides not showing a significant adjusted p-value. In this case, using Limma would have resulted in missing the majority of differentially modified peptides.

@@

The results of this experiment are summarized in Figure \ref{fig:data4_plots}. The experiment was modeled as a group comparison. In Figure \ref{fig:data4_venn_diagram} the number of significant modified peptides before and after adjustment is shown. We can see that more peptides became insignificant after adjustment than became significant. 3,236 modified peptides became insignificant, 1352 became significant, while 4,282 were significant in both models. For the peptides that became insignificant in the adjusted model, their change in abundance was mainly due to changes in global protein abundance. In contrast, for peptides that became significant after adjustment, their true abundance change was masked by underlying changes in the unmodified protein. Both of these issues are corrected in adjustment, and the true abundance change is shown. An additional question that must be addressed is if the decrease in significant peptides is due to the increased variance that comes from adjustment. This was tested by looking for modified peptides whose adjusted log fold change was within 10\% of the unadjusted log fold change but became insignificant after adjustment. In other words, the fold change was the same between models but variance increased. When this test was applied on this experiment, only one peptide became insignificant due to an increase in variance. Thus we can conclude that the drop off in significant peptides was due to changes in global protein abundance.

>>>>

Each simulation was generated with random biological variation. Specifically, let $Y_{ibc}$ denote the abundance of peptide $i$ in biological replicate $b$ of condition $c$. The simulated peptide abundance  $Z_{ibc}$ was generated as $Z_{ibc} = Y_{ibc} + \epsilon_{ibc}$ where $\epsilon_{ibc}\sim^{iid}N(0,\sigma^2)$. The simulations were generated with one of two values $\sigma^2 = \{.2, .3\}$, motivated by the biological experiments in this manuscript.
Further details are in Supplementary Sec. \ref{supp-sec:dataset1}.

@@

\subsection*{Dataset 5 : Mouse - Phosphorylation - 2mix-TMT time series}
\label{sec:exp_proc_dataset5}
In this experiment 8 biological replicates were allocated to 2 TMT mixtures in an unbalanced time course design. More details are in Supplementary Sec. \ref{supp-sec:shigella}.

@@

The proteome was quantified in 11-plex isobaric multiplexing via tandem mass tagging (TMT) in combination with LC-MS/MS. The six conditions were split between 11 channels, and as the result the experiment was unbalanced. Each mixture contained two replicates per early and late WT and KO conditions. Mixture one contained one replicate of uninfected WT and two replicates of uninfected KO. Mixture two contained one replicate of uninfected KO and two uninfected WT.

@@

In this experiment 8 biological replicates were split between 4 conditions in a balanced group comparison design. Further details in Suppler Sec \ref{supp-sec:usp30}.

@@

Figure \ref{fig:data-structure}(a) schematically illustrates this data structure for one protein in a simplified experiment with $I=2$ conditions. The data are collected in $J=2$ \todo{biological?} replicate runs. Each run represents an observation used to estimate the relative abundance of the PTM in condition $i$, $u_i^{\ast}$. The number of quantified features vary across replicate LC-MS/MS runs of the same sample, and across conditions. There may be feature outliers which could be caused by random variation or measurement error. Additionally, some spectral features can be missing.

The population parameter we are interested in estimating is the difference between the PTM in condition $1$ and condition $2$, denoted by $u_1^{\ast}$ and $u_2^{\ast}$ respectively. To calculate the quantity of interest, statistical analysis needs to summarize values in the table in Figure \ref{fig:data-structure}(a) using appropriate models. For condition $i$, the models provide estimates for the mean abundance, $\hat{u}_i^{\ast}$, and standard error, $SE(\hat{u}_i^{\ast})$. Once summarized, we can formulate the Null and Alternative Hypotheses as seen in Equation \ref{eq:conv_null_hyp}.

\begin{equation}
\begin{aligned}
H_{0}: \Delta = \mu_{1}^{\ast} - \mu_{2}^{\ast} = 0 \\
H_{a}: \Delta = \mu_{1}^{\ast} - \mu_{2}^{\ast} \neq 0
\end{aligned}
\label{eq:conv_null_hyp}
\end{equation}

However, these population parameters are inherently convoluted with changes in the unmodified protein. To account for this, we must estimate the unmodified protein in conditions $1$ and $2$, denoted by $\mu_1$ and $\mu_2$. We estimate these quantities in the same way as for the PTM, using their correspond features and observations, resulting in an estimated mean abundance, $\hat{u}_i$, and standard error, $SE(\hat{u}_i)$. These estimates are used to adjust the quantities estimated for the PTM, resulting in the new Null and Alternative Hypotheses seen in Equation \ref{eq:null_hyp}.

\begin{equation}
\begin{aligned}
H_{0}: \Delta = ( \mu_{1}^{\ast} - \mu_{1}) - ( \mu_{2}^{\ast} - \mu_{2} ) = 0 \\
H_{a}: \Delta = ( \mu_{1}^{\ast} - \mu_{1}) - ( \mu_{2}^{\ast} - \mu_{2} ) \neq 0
\end{aligned}
\label{eq:null_hyp}
\end{equation}

@@

\noindent {\bf Two-sample $t$-test} \todo{When describing the existing methods, could you refer to Fig 1? Here is the suggested order: (1) What is the null hypothesis? Do these methods test the same null hypothesis? (If not, it may be a good idea to add this null hypothesis to the figure for contrast) (2) What are the data (how summarized prior to the testing); (3) what is the model assumptions? (4) how inference is performed}

\todo{These sentences were combined from different places. Could you edit?}

Many investigations perform differential abundance analysis of PTMs using two-sample $t$-test or its extensions. While simple, these approaches do not fully account for the sources of variations, and are not directly applicable to experiments with complex designs, e.g., comparisons of multiple conditions, acquisition in multiple batches, etc. Additionally, while these approaches can be applied to experiments studying PTMs, there is not a self contained, straightforward implementation of the methods, making application challenging.

The approach generally takes as input intensities of individual features from modified peptides, or intensity ratios of modified and unmodified peptide features, and compares the mean abundance of a PTM site from one condition to another~\cite{Schwammle2015}\cite{THOMAS2020}. Two-sample $t$-test is based on the null hypothesis that there is no difference in mean PTM abundance between Conditions $1$ and $2$. The abundance in each run is taken as input and is often estimated by sum of peak intensities. For example, the log-abundance estimate for the PTM in Run $j$ of Condition $1$ is given by

\begin{equation}
\begin{aligned}
\log \left( \sum_{k=1}^{K} 2^{y_{1jk}^{\ast}} \right).
\end{aligned}
\label{eq:feature_estimation}
\end{equation}

For adjustment with respect to unmodified peptides, the estimate of PTM abundance is divided by the protein abundance estimate.
%which is equivalent to take the difference of their log abundances. 
For condition $1$ the quantity is denoted by $d_{1j}$ and is given by

\begin{equation}
\begin{aligned}
d_{1j}^{\ast} = \log \left( \sum_{k=1}^{K} 2^{y_{1jk}^{\ast}} \right) - \log \left( \sum_{l=1}^{L} 2^{y_{1jl}} \right).
\end{aligned}
\label{eq:adj_estimation}
\end{equation}

$t$-test then takes the adjusted PTM abundance of condition $1$, $d_{1j}$, and condition $2$, $d_{2j}$, for each run in $J$ as input to estimate the quantity of interest $\Delta$, as well as the standard error, $SE(\Delta)$. This results in the Null and Alternative Hypotheses show in Equation \ref{eq:ttest_null_hyp}, where $\hat{d}_{1+}^{\ast}$ represents all $J$ runs for condition $1$ and is the estimate of $(\mu_{1}^{\ast} - \mu_{1})$ seen in Equation \ref{eq:null_hyp}.

\begin{equation}
\begin{aligned}
H_{0}: \Delta = d_{1+}^{\ast} - d_{2+}^{\ast} = 0 \\
H_{a}: \Delta = d_{1+}^{\ast} - d_{2+}^{\ast} \neq 0
\end{aligned}
\label{eq:ttest_null_hyp}
\end{equation}

%Alternatively, run-level summarization can also be used. The difference between the means of PTM abundance in Conditions $i$ and $i'$ are estimated as
%\[
%\hat{\Delta} = \frac{1}{J} d_{i+}^{\ast} - \frac{1}{J} d_{i'+},
%\]
%where $d_{i+}^{\ast} = \sum_{j=1}^{J} d_{ij}^{\ast}$ and the test statistic for the $t$-test is given by $\hat{\Delta} / \mathrm{SE}(\hat{\Delta})$. The statistical significance of the difference is determined by comparing the test statistic against the $t$ distribution, with degrees of freedom $df=2J-2$ in balanced designs.

\medskip \noindent {\bf Limma} Modifications of $t$-test such as moderated $t$-test with Limma were also proposed~\cite{Ritchie_15a}\cite{Zhu}. Limma uses linear models to test the null hypothesis that there is no difference in mean PTM abundances between conditions. It leverages Empirical Bayes moderation to share pooled variance information across individual modification models and moderate the individual residual variances. Using a linear model allows Limma to share variance information across conditions, providing a more accurate estimate~\cite{Ritchie_15a}.

With respect to PTM analysis, feature level summarization and unmodified protein adjustment are performed in the same way as $t$-test, shown in Equations \ref{eq:feature_estimation} and \ref{eq:adj_estimation}. Limma fits a linear model using the adjusted summarized features, $d_{1+}^{\ast}$ and $d_{2+}^{\ast}$, to estimate the difference between condition $1$ and $2$, $\Delta$. This results in the same hypothesis as seen in Equation \ref{eq:ttest_null_hyp}. The variance for the estimation is moderated using information from the variance across all PTMs modeled. For the estimate of a single PTM $m$, $\Delta_m$, the variance is estimated as $s_{\Delta_m} = P(v_{\Delta_m}| v_\Theta)$ where $v_{\Delta_m}$ is the resulting variance when modeling PTM $m$ and $v_\Theta$ is the global variance across all PTMs.

\todo{Could you fine-tune, using the same order as above for Limma? I.e., null hypothesis, the data, the model, the inference. I think the text can be made more compact, and a bit more technical, by directly referring to either Fig 1 or the equations above}

@@

The proposed approach leverages the large amount of work done to create $MSstats$~\cite{Choi:2014}. $MSstats$ can analyze the PTM without accounting for convolution with the unmodified protein. In terms of Figure \ref{fig:data-structure}, this means we can either model the modified or unmodified features, but we cannot combine them as seen in part Figure \ref{fig:data-structure}(c). The only Null and Alternative Hypotheses we can address are those in Equation \ref{eq:conv_null_hyp}, without adjusting for unmodified protein abundance.

$MSstats$ takes as input a list of log-transformed intensities of spectral features which are used to characterize the identified protein. For each protein, the feature log-intensities are expressed using a linear mixed model in consideration of the effects of condition, run, feature and interaction between run and feature. The model parameters are estimated using a split-plot approach, where the feature log-intensities are first summarized into a single value per protein per run in the subplot model, and the protein-level summaries are then used for the inference of the protein abundance. This approach allows the method to be extended to cases where the experimental design is unbalanced and additional sources of variation are present, overcoming the limitations in the previous sections. In the protein-level summarization, Tukey's median polish (TMP), a simple and robust procedure, is applied to iteratively fit a two-way additive model with the effects of run and feature, which in turn summarizes the log-intensities for each protein~\cite{Tukey:1977}. After summarization, the inference of the protein abundance in each condition is carried through fitting a model based on the family of linear mixed-effects models \cite{Bolker2009} \cite{Faraway:2006}. We leverage this statistical modeling and quantification workflow and apply it to PTMs, in addition to unmodified proteins.

@@

\todo{Could you also refer to Figure 1? Essentially, MSstats only analyzes a part of the figure. Refer to the null hypothesis. Then describe the summarization (including AFT etc), and the model for he summarized data (the fact that different designs require different models - this is your argument for the limitation of the methods above.)}

@@

We propose a statistical method for detecting changes in PTMs. The approach removes confounding with the unmodified protein, while being applicable to experiments with complex designs. For simplicity, we discuss the method in the special case of a balanced design with multiple conditions and one technical replicate. However, in  practice the experimental design can be unbalanced and variation can come from multiple sources, as shown in the next section.

The proposed approach first applies the methods used in $MSstats$ to both the PTM and unmodified protein separately. This includes run-level feature summarization using Tukey's Median Polish and fitting a model based on the family of linear mixed-effects models. After a model is fit for the both the PTM and unmodified protein, the modeling results are then combined to estimate the quantity of interest.

The null hypothesis states that there is no difference in log-abundance of the PTM site between conditions, adjusted with respect to protein abundance. Relating to the example in Figure \ref{fig:data-structure}, the expected value of log-abundance of the PTM and unmodified protein in Condition $1$ are denoted by $\mu_{1}^{\ast}$ and $\mu_{1}$ respectively. The feature and run level summarization for Condition $1$ is denoted as $\hat{y}_{1+}^{\ast}$ for the modified peptide and $\hat{y}_{1+}$ for the unmodified peptide. The Null and Alternative Hypotheses for Conditions $1$ and $2$ are shown in Equation \ref{eq:prop_null_hyp}. This is in contrast to the hypotheses show in Equation \ref{eq:null_hyp}, where the PTM was adjusted before modeling.

\begin{equation}
\begin{aligned}
H_{0}: \Delta = ( \mu_{1}^{\ast} - \mu_{2}^{\ast} ) - ( \mu_{1} - \mu_{2} ) = 0 \\
H_{a}: \Delta = ( \mu_{1}^{\ast} - \mu_{2}^{\ast} ) - ( \mu_{1} - \mu_{2} ) \neq 0
\end{aligned}
\label{eq:prop_null_hyp}
\end{equation}

Specifically, the adjusted difference is given by the difference in log-abundance of the PTM site, subtracted by the difference in log-abundance of the underlying protein, which is equivalent to the log of the ratio of PTM abundance difference to protein abundance difference, in a balanced case.

\begin{equation}
\hat{\Delta} = \left[ \frac{1}{J} \left( \hat{y}_{1+}^{\ast} - \hat{y}_{2+}^{\ast} \right) \right] - \left[ \frac{1}{J} \left( \hat{y}_{1+} - \hat{y}_{2+} \right) \right]
\label{eq:prop_fc}
\end{equation}

The estimate of the adjusted difference and the standard error (SE) of the estimate are obtained by combining the difference estimates and the associated SEs from both counterparts. The equation accounts for the increased uncertainty when incorporating more information from the unmodified protein. This may lead to an increased SE, when compared to modeling the PTM alone.

\begin{equation}
SE(\hat{\Delta}) = \left[ \left( \hat{\sigma}_{\gamma^{\ast}}^{2} + \hat{\sigma}_{\gamma}^{2} \right) \right]^{1/2}
\label{eq:prop_se}
\end{equation}

@@

\begin{equation}
\left( \hat{\sigma}_{\gamma^{\ast}}^{2} + \hat{\sigma}_{\gamma}^{2} \right)^2 \bigg/
\left( \frac{\hat{\sigma}_{\gamma^{\ast}}^{4}}{\mathrm{df}(\gamma^{\ast})} + \frac{\hat{\sigma}_{\gamma}^{4}}{ \mathrm{df}(\gamma)} \right)
\label{eq:prop_df}
\end{equation}

@@

\todo{Put it in the context of the figure again, and also in the context of the existing methods in Background - how is it similar to the existing methods, and how different}

@@

The proposed approach can be extended to complex designs, including experiments with additional sources of variation and unbalanced designs. In this case our null hypothesis in Equation \ref{eq:prop_null_hyp} is unchanged.

Model inference is done in the same way to how MSstats~\cite{Choi:2014} and MSstatsTMT~\cite{Huang:2020} target the global protein. In cases where the design is unbalanced, a linear mixed effects model is fit, which takes into account all potential sources of variation, and restricted maximum likelihood (RML) is used to estimate the parameters of each model. Once the parameters are estimated, we can combine the models using modified versions of Equations \ref{eq:prop_fc}, \ref{eq:prop_se}, and \ref{eq:prop_df}.

The log-fold change in the adjusted PTM abundance, $\Delta$, is now estimated by RML, and Equation \ref{eq:prop_fc} becomes 
\begin{equation}
\hat{\Delta} = (\hat{\mu}_{RML_i}^{\ast} - \hat{\mu}_{RML_{i'}}^{\ast}) - (\hat{\mu}_{RML_i} - \hat{\mu}_{RML_{i'}})
\end{equation}

@@

The proposed methods are implemented in the open source R package $MSstatsPTM$, available on Bioconductor and Github. An overview of the steps of the package are illustrated in Figure \ref{fig:msstatsptm_design}. $MSstatsPTM$ includes converters for multiple spectral processing tools, including MaxQuant, Progenesis, and Spectronaut. The converters take as input the raw data from the tool, identify the modification site for modified peptide, and put the data into the correct format for analysis in $MSstatsPTM$. Conversion is done separately for the modified and global profiling runs. If the global profiling run is not available, the package can still analyze the modified run, but will do so without adjusting for changes in unmodified protein abundance. Specifically, if a global profiling run is not available the hypotheses in Equation \ref{eq:conv_null_hyp} will be used; which amounts to running the methods seen in $MSstats$ on modified peptide level data \cite{Choi:2014} \cite{Huang:2020}.

After using the converters, the next step is peptide/protein summarization and missing value imputation. For the modified run, the package summarizes features which include the same modification together. Features with multiple modification sites are not included with single site features and are summarized separately. For the global profiling run all unmodified features from the same protein are summarized together, up to the protein level. Additionally, the summarization includes global median normalization and normalizes between MS runs. The package can optionally impute missing values using an AFT model.

@@

The package relies on functionalities from the R packages $MSstats$ \cite{Choi:2014} and $MSstatsTMT$ \cite{Huang:2020}. The statistical modeling relies on the functionality from the R packages $lme4$ \cite{Bates2015} and $lmerTest$ \cite{Kuznetsova2017}.

@@

In this simulation the proposed method outperformed $t$-test and Limma. Changes in unmodified protein level still needed to be accounted for to control the FDR. Once controlling for changes in the unmodified protein, the proposed method outperformed the other methods, as seen in Figure \ref{fig:sim2_acc}. The proposed method calibrated model accuracy and recall rate well, even when the number of replicates were low. Additionally, when comparing the fold change estimation across all modified peptides, the proposed method showed a tighter distribution of estimated fold changes around the true fold change, as compared to $t$-test and Limma. Specifically, the inner quartile range (IQR) of the estimated fold change for the proposed method was on average 10.4\% smaller than $t$-test and 21.8\% smaller than Limma. This fold change comparison can be seen in Supplementary Figure \ref{supp-fig:fc_boxplot}.

@@

In this experiment all models incorrectly estimated the fold change of the modified spike-in peptides before adjusting for changes in global protein abundance. After adjustment, the spike-in peptides' fold change was generally in line with expectation in all methods, however the distribution of estimated fold changes was visibly wider (Figure \ref{fig:spikein_boxplot}). Of the three approaches, the proposed method showed the tightest distribution around the true log fold change. Comparing the IQR of the spike-in peptide's log fold change, the proposed method's IQR was 36.78\% smaller than Limma's and 32.98\% smaller than $t$-test's. This means that the proposed method's fold change estimate was more consistently correct when compared to Limma and $t$-test.

In Figure \ref{fig:spikein_prop_volcano} we can clearly see the red labeled spike-in peptides do not follow the expected fold change before adjusting for changes in global protein abundance. After adjustment, the estimated fold change was more in line with expectation. Additionally, the background peptides serving as the null model show many false positives before adjustment was made, however after adjustment the number of false positives decreased significantly. Specifically, for the proposed method, the number of false positives went from 20.88\% to 1.84\% after adjustment was applied. Using both the proposed method and Limma, the red labeled spike-in peptides were more inline with expected fold change after the adjustment was applied. However, Limma resulted in the majority of the differential spike-in peptides having an insignificant adjusted p-value. In this case, using Limma would have resulted in missing the majority of differentially modified peptides.

@@

The results of this experiment are summarized in Figure \ref{fig:data4_plots}. In Figure \ref{fig:data4_venn_diagram} the number of significant modified peptides before and after adjustment is shown. We can see that more peptides became insignificant after adjustment than became significant. 3,236 modified peptides became insignificant, 1352 became significant, while 4,282 were significant in both models. For the peptides that became insignificant in the adjusted model, their change in abundance was mainly due to changes in global protein abundance. In contrast, for peptides that became significant after adjustment, their true abundance change was masked by underlying changes in the unmodified protein. Both of these issues are corrected in adjustment, and the true abundance change is shown. An additional question that must be addressed is if the decrease in significant peptides is due to the increased variance that comes from adjustment. This was tested by looking for modified peptides whose adjusted log fold change was within 10\% of the unadjusted log fold change but became insignificant after adjustment. In other words, the fold change was the same between models but variance increased. When this test was applied on this experiment, only one peptide became insignificant due to an increase in variance. Thus we can conclude that the drop off in significant peptides was due to changes in global protein abundance.

In Figure \ref{fig:data4_profile_plot}, the modification of protein $GSDMD$ at site $K62$ showed the advantage of the proposed method. The modified peptide originally showed no abundance change between the infected 1 hour, 4 hour, and 6 hour conditions. This was contrasted with a strong negative change in the global profiling run between the same conditions. Looking at the Dox4hr vs Dox1hr conditions and modeling the modified peptide without adjusting for changes in the global profiling, the fold change was $-.501$ and the adjusted p-value was insignificant at $.0644$. After adjusting for changes in the global protein abundance, the fold change was much higher, $2.79$, and the adjusted p-value became significant, $5.25e-8$. In this case the effect of the modified peptide was strongly confounded with changes in the global protein. The proposed method allowed us to remove this confounding and estimate the true effect.
